Open archaeology, open source?

### 2024-04-15 analysis/_02-background.qmd
Add archaeological examples

>>>>

As a concrete example of how git and GitHub is used by archaeologists, we can take <https://open-archaeo.info> itself.
The website at that address was at the time of writing generated from a *git* source repository containing the data on the individual entries, a set of HTML templates, and some R scripts that translate between them.
A copy of this repository can be downloaded and worked on locally by anyone, who will have access not just to the current state of the source code but its full history through the git *version control* software.
The authoritative version is hosted on *GitHub* at <https://github.com/zackbatist/open-archaeo>.
The GitHub URL indicates the primary *maintainer* of the repository, 'zackbatist' – that is, one of us (Zack) – who created and has ultimate control over it.
However, various other *contributors* (such as the other of us, Joe) have added entries, corrected entries, or added functionality to the website using the *fork* and *pull request* features of GitHub.
The basic unit of this type of contribution is the *commit*, which is a discrete set of changes (e.g. adding an entry) associated with one contributor at a single point of time.
Others have contributed by raising *issues* describing problems with or suggestions for the project, leaving *comments* on these issues, or more loosely by using GitHub's social media features (*stars* and *following*).
The full history of all these types of contributions to open-archaeo can be accessed through GitHub – or, as we use here, its programmatic API.
This basic workflow is the same whether the project in question is a document like open-archaeo, or a piece of software, or a research paper; though actual patterns of collaboration vary markedly, as we will see.

### 2024-04-15 analysis/_04-open_archaeology.qmd
Add archaeological examples

In our breakdown of open-archaeo by category (see @tbl-categories), we demonstrate the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language or a plugin for an application.
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

@@

Archaeologists have been using git from the early 2000s, before GitHub was launched,^[Commits dated to before GitHub's launch in 2008 come from repositories that were created earlier and later imported to GitHub's platform. We have excluded commits that have obviously erroneous dates, e.g. `2001-01-01` in projects started in the mid–late 2010s.].
but it saw a marked increase in popularity c. 2014--2015.
From that point and until recently there was an exponential uptake of GitHub by archaeologists,
but while we were preparing this paper (c. 2022) the first signs of a diminishing growth have appeared,
with the number of cumulative commits continuing to rise but the number of repositories hitting a sharp plateau.
This could be explained by market exhaustion, a shift in emphasis to maintaining existing code and working on established projects, and/or growing doubts about the appropriateness and sustainability of GitHub following its acquisition by Microsoft [@kansa2022b].^[Concerns we very much share.]

>>>>

In our breakdown of open-archaeo by category (see @tbl-categories), we demonstrate the pervasiveness of various development models, and the requisite technical capabilities that developers assume users hold.
Most resources (`r percent(p_platform)`) included in open-archaeo are designed to be used atop an existing "platform" -- for example a package that extends a programming language (e.g. radiocarbon calibration is implemented in R in the package 'rcarbon' or Python in the package 'iosacal') or a plugin for an application (e.g. 'ArchaeoAstroInsight' adds tools for measuring astronomical alignments to QGIS).
Essentially such projects create additional functions within the base platform that are useful for archaeological purposes.
Others create standalone software (`r percent(p_standalone)`) that can be run independently of such platforms, for example desktop or web apps.
A significant number of projects also comprise of datasets (`r percent(p_datasets)`) and non-packaged code snippets (`r percent(p_scripts)`) that have been made available for general use.

@@

Archaeologists have been using git from the early 2000s, before GitHub was launched,^[Commits dated to before GitHub's launch in 2008 come from repositories that were created earlier and later imported to GitHub's platform. We have excluded commits that have obviously erroneous dates, e.g. `2001-01-01` in projects started in the mid–late 2010s.].
but it saw a marked increase in popularity c. 2014--2015.
From that point and until recently there was an exponential uptake of GitHub by archaeologists,
but while we were preparing this paper (c. 2022) the first signs of a diminishing growth have appeared,
with the number of cumulative commits continuing to rise but the number of repositories hitting a sharp plateau.
This could be explained by market exhaustion, a shift in emphasis to maintaining existing code and working on established projects, and/or growing doubts about the appropriateness and sustainability of GitHub following its acquisition by Microsoft [@kansa2022].^[Concerns we very much share.]
